
This markdown file contains visualization work related to the JOUR328G High School NIL Policy Reporting Project.

```{r}

library(tidyverse)
library(ggalt)
library(ggplot2)
library(readr)
library(usmap)
library(maps)
library(dplyr)
library(viridis)
library(scales)
library(gt)
library(ggrepel)

```

Load in the binary answer file.

```{r}

hs_nil <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/nil_clean.csv")
hs_nil
```

Categories included so far are: 
- State Policy: Y/N
- Year Enacted:
- Did we get a response from an admin?: Y/N
- Parent or guardian permission required for entering a deal?: Y/N
- Are boosters or collectives permitted?: Y/N
- Requirement to notify school officials?: Y/N
- Is there an applicable state law?: Y/N
- Are there category restrictions?: Y/N

Points of clarification for us to work through together as a class:
- Our question about usage of school logos, names or uniforms: Are there instances where a policy specifies that athletes can use two of the three or are all three typically grouped together?
School property, school-sanctioned events

- Is there a discrepancy between a policy's enacted date versus the date it went into effect? Which do we prefer as a class?


- Prohibited Categories: How can we better group some of our categories for visualization? Does it make sense to create a "super category" for tobacco, nicotine, vaping and regulated/controlled substances? How can we also group illegal substances or gaming and gambling?
- Groupings ready for viz next week.

- NIL Payments: Can we identify whether schools are tracking individual payments or total payments over a period of time? Are there compensation limits in either of those instances (i.e., can only earn X dollars over one school year vs X dollars in one payment). Which do these policies focus on, if specified?
- Focus timing for the reporting of these deals, not so much the payment values

- NIL Violations and Accountability: Is there an abdication of adult/school responsibility in states where no parental guidance or school notice is required? Who is held accountable in those cases?  
Highlight Alaska, but also any other states that have a three-strike model. South Carolina has a three-strike policy for athletes.

- Officiating athletic contests is a potential unique reporting line: Is Kansas the only state to make reference of student-athletes being permitted to be compensated as referees, but not as coaches? Why make that distinction in the first place? What's the state's policy on appearances for NIL and would these not fall under that? 


Regarding all of these points and of the visualizations below: How are we going to describe this when we go to write it? If we were to write a paragraph about this question, what would it look like?

------------
Map visualization of state policies

```{r}

us_map <- map_data("state")

nil_states <- hs_nil |>
  mutate(state = tolower(state))

us_map_data <- us_map("states") |>
  mutate(full = tolower(full))  

nil_usmap_data <- us_map_data |>
  left_join(nil_states, by = c("full" = "state"))

nil_usmap_data <- nil_usmap_data |>
  mutate(state_policy = case_when(
    full == "north carolina" ~ "Y",
    full == "ohio" ~ "N",
    full == "indiana" ~ "N",
    full == "tennessee" ~ "Y",
    full == "utah" ~ "Y",
    full == "vermont" ~ "Y",
    TRUE ~ state_policy 
  ))


```

```{r}

plot_usmap(data = nil_usmap_data, values = "state_policy", regions = "states") +
  scale_fill_manual(values = c("Y" = "blue", "N" = "red", "Limited" = "yellow"),
                    labels = c("Y" = "Yes", "N" = "No", "Limited" = "Limited")) +
  labs(title = "NIL Policy Status by U.S. State", fill = "Policy Status") +
  theme(panel.background = element_blank())

```
Notes:

----------------
Map of the South Region Based on NIL Policy Adoption.

```{r}

plot_usmap(data = nil_usmap_data, values = "state_policy", include = .south_region, exclude = c("VA", "WV", "MD", "DE", "DC", "KY"), regions = "states", labels = TRUE) +
  scale_fill_manual(
    values = c("Y" = "blue", "N" = "red", "Limited" = "yellow"),
    labels = c("Y" = "Yes", "N" = "No", "Limited" = "Limited")
  ) +
  labs(title = "The South Lags in HS NIL Policy Adoption", fill = "Policy Status") +
  theme(panel.background = element_blank())

```

What does this show us? Why might there be such a high concentration of states in this region that have declined to enact an NIL policy for high schoolers? Also, can we figure out approximately how many high school athletes live in this region that are impacted by these decisions?

Notes: (Fixed)

---------------
Years enacted: 

```{r}

yearly_enactments <- nil_states |>
  filter(!is.na(year_enacted)) |>
  group_by(year_enacted) |>
  summarise(count = n())

```

```{r}

max_count <- max(yearly_enactments$count, na.rm = TRUE)

ggplot(yearly_enactments, aes(x = factor(year_enacted), y = count)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Number of States Enacting NIL Policies by Year",
       x = "Year Enacted",
       y = "Number of States") +
  scale_y_continuous(breaks = seq(0, max_count, by = 1)) +  
  theme_minimal()

```
Looking at these results, let's see what we can find out about that jump in 2022. Which states set the tone in 2021 and what do we think was the catalyst in 2022?
Notes:


```{r}
nil_usmap_data <- nil_usmap_data |>
  mutate(
    year_enacted = as.character(year_enacted),
    year_enacted_override = ifelse(full == "north carolina", "2024", year_enacted)  # Overrides North Carolina to match the "2024" color
  )

# Define colors without a separate "NC" entry
year_colors <- c("2021" = "yellow", "2022" = "darkgreen", "2023" = "purple", "2024" = "blue", "No NIL" = "grey")

# Plot the map with customized color for North Carolina
plot_usmap(data = nil_usmap_data, values = "year_enacted_override", regions = "states") +
  scale_fill_manual(
    name = "Year Enacted", 
    values = year_colors,
    na.translate = FALSE 
  ) +
  labs(
    title = "NIL Policy Adoption by Year", 
    fill = "Year Enacted", 
    subtitle = "Preliminary visualization"
  ) +
  theme(panel.background = element_blank())

```

Notes:

Proposed but denied is a potential graphic.


-----------------------
Admin response: These states played ball. 

```{r}

nil_usmap_data <- nil_usmap_data |>
  mutate(
    employee_response = ifelse(full == "north carolina", "NC", employee_response)
  )

plot_usmap(data = nil_usmap_data, values = "employee_response", regions = "states") +
  scale_fill_manual(
    values = c("Y" = "blue", "N" = "red"),  
    na.value = "lightgray"
  ) +
  labs(
    title = "These States Played Ball", 
    subtitle = "Visual distribution of states that provided a policy response to our reporters.", 
    fill = "Employee Response"
  ) +
  theme(panel.background = element_blank()) +
  
  geom_sf(data = subset(nil_usmap_data, full == "north carolina"), fill = "blue", color = NA, show.legend = FALSE)

```
Notes:

New Mexico -- response received, but didn't "feel comfortable" answering questions. (Fixed)

------------------------
Parent/Guardian Involvement and School Notification:

```{r}

pg_sn <- nil_usmap_data |>
  mutate(
    parent_notification_combo = case_when(
      parent_guardian == "Y" & notification == "Y" ~ "Y_Y",
      parent_guardian == "Y" & notification == "N" ~ "Y_N",
      parent_guardian == "N" & notification == "Y" ~ "N_Y",
      parent_guardian == "N" & notification == "N" ~ "N_N",
      parent_guardian == "No NIL" ~ "No NIL",   
      TRUE ~ NA_character_  
    )
  )

combo_colors <- c("Y_Y" = "blue", "Y_N" = "purple", "N_Y" = "yellow", "N_N" = "red", "No NIL" = "cyan")

```

```{r}

pg_sn <- nil_usmap_data |>
  mutate(
    parent_notification_combo = case_when(
      parent_guardian == "Y" & notification == "Y" ~ "Y_Y",
      parent_guardian == "Y" & notification == "N" ~ "Y_N",
      parent_guardian == "N" & notification == "Y" ~ "N_Y",
      parent_guardian == "N" & notification == "N" ~ "N_N",
      parent_guardian == "No NIL" ~ "No NIL",   
      TRUE ~ "Not Specified"   
    ),
    parent_notification_combo = addNA(parent_notification_combo)  
  )

combo_colors <- c("Y_Y" = "blue", "Y_N" = "purple", "N_Y" = "yellow", "N_N" = "red", "No NIL" = "cyan", "Not Specified" = "lightgray")
```

```{r}

pg_sn |>
  group_by(parent_notification_combo) |>
  summarise(count = n())

```



```{r}

plot_usmap(data = pg_sn, values = "parent_notification_combo", regions = "states") +
  scale_fill_manual(
    name = "Requirement",
    values = combo_colors,
    labels = c(
      "Y_Y" = "Parent/Guardian and School Notif. Required",
      "Y_N" = "Parent/Guardian Required Only",
      "N_Y" = "School Notif. Required Only",
      "N_N" = "Neither Required",
      "No NIL" = "No NIL",
      "NA" = "Not Specified" 
    )
  ) +
  labs(
    title = "Parent/Guardian and School Notification Requirements for NIL Policies", 
    fill = "Requirement"
  ) +
  theme(
    panel.background = element_blank(),
    legend.position = "left",               
    legend.justification = "center",       
    plot.margin = margin(10, 10, 10, 30),   
    plot.title.position = "plot",          
    plot.title = element_text(hjust = 0)   
  )

```

Notes:

- The number of states: Do a cartogram, the number of states (waffle chart? these number of states have this, emphasize the difference in choices here. Visualize the fact that very few states have a reporting requirement.)

------------------------
What else do we want to see? What direction do we want to take with our reporting? What changes can we make to the Google Form to continue refining this data?

-------------------------------------------------------
###Updates for 11/11/24 Class:

Feedback from the students on the previous visualizations has been implemented.

- Reporting Ideas: 
1. Three-Strike States: Pennsylvania (2022), South Carolina (2024), Florida (2024)
What can we identify regarding the differences between these policies? What prompted them to institute a three-strike policy? Why are they the only three to implement a policy like this? What did the more recent policies adopt from Pennsylvania's policy?

2. Limited Approval States: Mississippi (2024), Missouri (2023), Arkansas (2023), Nevada (2022)
Mississippi allows it for students committed to any collegiate institution, Missouri and Arkansas allow it for students committed to in-state collegiate institutions and Nevada only allows it for unsanctioned sports.

3. NFTs? Vermont, Maryland and Massachusetts all specifically cite NFTs as a permitted NIL category. 
Are high schoolers taking them up on that? Why bother explicitly including such a niche category? 


What I'm working on right now: Creating a separate data frame for violations. I've created a data frame for approved and prohibited NIL categories and will begin visualizing that.

Visualizations:

Map of US States filled in based on the number of prohibited categories

---------
Stacked Bar Chart with the Count of Prohibited NIL Policy Categories

*DON'T USE THIS VISUALIZATION. KEEPING IN DOC FOR SHOWING WORK
```{r}

perm_cat <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/nil_permitted_categories.csv")
proh_cat <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/nil_prohibited_categories.csv")

```

```{r}

#nil_proh <- proh_cat_long |>
#  filter(status %in% c("Y", "No NIL", "Not Specified")) |>
#  group_by(category, status) %>%
#  summarise(total = n(), .groups = 'drop')

#nil_proh$category <- factor(nil_summary$category, levels = c("adult_ent", "alcohol", "controlled_sub", 
                                                             #   "gambling", "prescriptions", "smoke_vape", 
                                                             #   "weapons", "other"))

#nil_proh <- proh_cat_long |>
#  filter(status %in% c("Y", "No NIL", "Not Specified")) |>
#  group_by(category, status) %>%
#  summarise(total = n(), .groups = 'drop')


# nil_proh$category <- factor(nil_proh$category, levels = c("adult_ent", "alcohol", "controlled_sub", 
                                                              #  "gambling", "prescriptions", "smoke_vape", 
                                                              #  "weapons", "other"))

#ggplot(nil_proh, aes(x = category, y = total, fill = status)) +
#  geom_bar(stat = "identity") +
#  labs(title = "Totals for NIL Prohibited Categories",
#       x = "Category", y = "Total Count",
#       fill = "Status") +
#  scale_x_discrete(labels = c(
#    "adult_ent" = "Adult Ent.",
#    "alcohol" = "Alcohol",
#    "controlled_sub" = "Controlled Sub.",
#    "gambling" = "Gambling",
#    "prescriptions" = "Prescriptions",
#    "smoke_vape" = "Smoke-related",
#    "weapons" = "Weapons",
#    "other" = "Other"
#  )) +
#  scale_fill_manual(
#    values = c("Y" = "red", "No NIL" = "cyan", "Not Specified" = "grey"),
#    labels = c("Y" = "Prohibited", "No NIL" = "No NIL", "Not Specified" = "Not Specified")
#  ) +
#  scale_y_continuous(breaks = seq(0, 55, by = 5), limits = c(0, 55)) +
#  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
Notes:
-----

*DON'T USE THIS VISUALIZATION. KEEPING IN DOC FOR SHOWING WORK
```{r}
#perm_cat_long <- perm_cat |>
#  pivot_longer(cols = -state, names_to = "category", values_to = "status")

#perm_summary <- perm_cat_long |>
#  filter(status %in% c("Y", "No NIL", "Not Specified", "N")) |>
#  group_by(category, status) |>
 # summarise(total = n(), .groups = 'drop')

#perm_summary$category <- factor(perm_summary$category, levels = c("endorsements", "appearances", 
                                                                 # "social_media", "product_promotion", "nfts"))

# ggplot(perm_summary, aes(x = category, y = total, fill = status)) +
#  geom_bar(stat = "identity") +
#  labs(title = "Totals for NIL Permitted Categories",
#       x = "Category", y = "Total Count",
#       fill = "Status") +
#  scale_x_discrete(labels = c(
#    "endorsements" = "Endorsements",
#    "appearances" = "Appearances",
#    "social_media" = "Social Media",
#    "product_promotion" = "Product Promotion",
#    "nfts" = "NFTs"
#  )) +
#  scale_fill_manual(
 #   values = c("Y" = "red", "No NIL" = "cyan", "Not Specified" = "grey", "N" = "purple"),
 #   labels = c("Y" = "Permitted", "No NIL" = "No NIL", "Not Specified" = "Not Specified", 
#              "N" = "Prohibited")
#  ) +
#  scale_y_continuous(breaks = seq(0, 55, by = 5), limits = c(0, 55)) +
#  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
#fix(perm_cat)
```

-------------

*DON'T USE THIS VISUALIZATION. KEEPING IN DOC FOR SHOWING WORK
Map Visualization of Number of Prohibited Categories by State


```{r}
#proh_cat <- proh_cat |>
#  mutate(state = tolower(state))

#map_data <- nil_usmap_data |>
#  left_join(proh_cat |> 
#    select(state, catcount), by = c("full" = "state")) 

#  map_data <- map_data |>
#  mutate(catcount = factor(catcount),
#  catcount = ifelse(full == "north carolina", 7, catcount),
#         catcount = factor(catcount))

```

```{r}

#plot_usmap(data = map_data, values = "catcount", regions = "states") +
#  scale_fill_manual(values = c(
#    "0" = "lightblue",
#    "2" = "skyblue",
#    "3" = "dodgerblue",
#    "4" = "blue",
#    "5" = "navy",
#    "6" = "darkblue",
#    "7" = "purple"
#  ), na.value = "grey80") +
#  labs(title = "Number of Prohibited NIL Categories by U.S. State", fill = "Prohibited\nCategories") +
#  theme(panel.background = element_blank())

```
Notes: 
- Need to specify which states have no NIL policies. Also, the students should double-check their states.

----------------

Visualization of State Violations (Bar Chart?)
Ideas: 
bar chart for number of states that result in students losing eligibility

---------
Notes: 
Footnote for "No Language" or "Vague"

---------------------------------------
### 11/18/24 Class Update:

First off, here's an interesting finding from my research this week:
Virginia's VHSL has a free NIL Policy Digital Course. 
The description: "The purpose of the VHSL NIL Policy digital course is to help students and their families to understand Name, Image, and Likeness. The course includes 8 video chapters – and downloadable content – that last 30 minutes in total." 
Link: https://www.vhsl.org/vhsl-nil-policy-digital-course/ 

Do any other states offer this level of education for their high school NIL policies? Might be worth looking into. Also, whoever has Virginia might want to enroll to let us know what insights we can glean from this.

This week I'm visualizing the potential impact of these high school policies based on data provided by the National Federation of State High School Associations, an advocacy nonprofit for high school athletics. For the purposes of this week's update, I'm using the numbers shown in the organization's 2023-2024 High School Athletics Participation Survey, which shows the number of high school athletes across all 50 states and Washington, D.C., by year, sport and gender.

Here's a link to the survey archives: https://www.nfhs.org/sports-resource-content/high-school-participation-survey-archive/

Here's a link to the most recent survey that I'm using:  https://www.nfhs.org/media/7213111/2023-24-nfhs-participation-survey-full.pdf

Visualization #1: High School Participation by State for 2023-2024

```{r}

tot_hs <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/tot_hs_participation.csv")

tot_hs <- tot_hs |>
  mutate(state = tolower(State)) |> 
  select(state, Total)

plot_usmap(data = tot_hs, values = "Total", regions = "states") +
  scale_fill_viridis(
    option = "plasma",  # Other options: "plasma," "viridis", "magma", "cividis"
    name = "Total Athletes",
    label = scales::comma,
    direction = -1  
  ) +
  theme_minimal() +
  labs(
    title = "Total High School Athletes by State (2023-24)",
    subtitle = "Data includes boys and girls participation rates",
    caption = "Data: NFHS"
  )

```

Notes:



---------

Visualization #2: High School Athlete Participation by Year

```{r}

part_by_year <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/hs_part_year.csv")

pby_15 <- tail(part_by_year, 15)

ggplot(data = part_by_year, aes(x = Year)) + 
  geom_line(aes(y = Total, color = "Total"), size = 1) +
  geom_line(aes(y = `Boys Participants`, color = "Boys"), size = 1) +
  geom_line(aes(y = `Girls Participants`, color = "Girls"), size = 1) +
  scale_color_manual(
    values = c("Total" = "darkgreen", "Boys" = "blue", "Girls" = "pink")
  ) +
  scale_y_continuous(labels = comma) +  
  labs(
    title = "High School Sports Participation Since 1971",
    subtitle = "Post-Covid jump in participation, coupled with high school NIL",
    x = "Year",
    y = "Participants",
    caption = "Data: NFHS",
    color = "Legend"
  ) +
  theme_minimal()

```
Notes:
Is participation skewed by mandatory participation among associations?


```{r}
pby_10 <- pby_15 |>
  arrange(Year) |>  
  slice(-1:-5) |>
  arrange(desc(Year))

pby_10 |> 
  gt() |>
  tab_header(
    title = "High School Sports Participation (Last 10 Years)",
    subtitle = "Comparison of boys, girls and total participants") |>
  fmt_number(
    columns = vars(`Boys Participants`, `Girls Participants`, Total),
    decimals = 0) |>
  cols_label(
    Year = "Year",
    `Boys Participants` = "Boys",
    `Girls Participants` = "Girls",
    Total = "Total") |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())) |>
  tab_source_note(
    source_note = "Data Source: NFHS") |>
  cols_align(
    align = "center",
    columns = everything())

```
Notes:




```{r}
ggplot(data = pby_10, aes(x = Year)) + 
  geom_line(aes(y = Total, color = "Total"), size = 1) +
  scale_color_manual(
    values = c("Total" = "darkgreen")) +
  scale_y_continuous(labels = scales::comma) +  
  scale_x_continuous(
    breaks = seq(min(pby_10$Year), max(pby_10$Year), by = 2)
  ) +
  labs(
    title = "High School Sports Participation Reaches Historical Peak Post-Pandemic",
    x = "Year",
    y = "Participants",
    caption = "Data: NFHS",
    color = "Legend"
  ) +
  theme_minimal()
```

Notes:


--------
Visualization #3: Participation by Sport
The sports pulled from the survey are ones that are accounted for in at least 40 states. I did not include male figures for softball or female figures for baseball as they did not meet the 40 state reporting threshold.

```{r}

sport_part <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/part_by_sport.csv")

```
```{r}

sport_part_long <- sport_part |>
  pivot_longer(
    cols = c(boys, girls),  
    names_to = "Gender",
    values_to = "Participants") |>
  mutate(Participants = as.numeric(gsub(",", "", Participants))) 

ggplot(sport_part_long, aes(x = sport, y = Participants, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c("boys" = "blue", "girls" = "pink")  
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "High School Sports Participation by Sport and Gender",
    x = "Sport",
    y = "Participants",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Notes:
How can we add context for this chart through reporting (i.e., Do we have a sense for the distribution of HS NIL deals by sport?)

Girls playing 11 on 11.

Here's the same info in a table format:
```{r}

sport_part_gt <- sport_part |>
  mutate(
    boys = as.numeric(gsub(",", "", boys)),
    girls = as.numeric(gsub(",", "", girls)),
    total = as.numeric(gsub(",", "", total))) |>
  arrange(desc(total))

sport_part_gt |>
  gt() |>
  tab_header(
    title = "High School Sports Participation by Sport",
    subtitle = "Boys, girls and total participants from programs reported by at least 40 states.") |>
  fmt_number(
    columns = c(boys, girls, total),
    decimals = 0) |>
  cols_label(
    sport = "Sport",
    boys = "Boys",
    girls = "Girls",
    total = "Total") |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())) |>
  cols_align(
    align = "center",
    columns = everything()) |>
  tab_source_note(
    source_note = "Data Source: NFHS")

```

Notes:



------------

Visualization #4: Comparison between boys and girls schools and sports.

```{r}
#code for creating a csv from pdf data

#boys <- tribble(
#  ~Sport, ~Schools, ~Participants,
#  "Basketball", 18587, 536668,
#  "Track and Field – Outdoor", 17274, 625333,
#  "Baseball", 16146, 471701,
#  "Cross Country", 15511, 239381,
#  "Golf", 14055, 155174,
#  "Football – 11-Player", 13788, 1031508,
#  "Soccer", 12754, 467483,
#  "Wrestling", 11149, 291874,
#  "Tennis", 10175, 157835,
#  "Swimming & Diving", 7855, 116799)

#girls <- tribble(
#  ~Sport, ~Schools, ~Participants,
#  "Basketball", 18090, 367284,
#  "Track and Field – Outdoor", 17267, 506015,
#  "Volleyball", 16849, 479125,
#  "Softball – Fast Pitch", 15635, 345451,
#  "Cross Country", 15228, 192969,
# "Soccer", 12307, 383895,
#  "Golf", 11089, 138174,
#  "Tennis", 10645, 195766,
#  "Swimming & Diving", 8133, 138174,
#  "Competitive Spirit", 7827, 101204)

#boys <- boys |> mutate(Gender = "Boys")
#girls <- girls |> mutate(Gender = "Girls")

#combined <- bind_rows(boys, girls)

#write_csv(combined, "top_sports_schools.csv")

sports_and_schools <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/hs_nil/top_sports_schools.csv")

```

```{r}

ggplot(sports_and_schools, aes(
  x = Schools,
  y = Participants,
  size = Participants,
  label = Sport,
  color = Gender
)) +
  geom_point(alpha = 0.6) +  
  geom_text_repel(size = 3, max.overlaps = 10) +  
  scale_size_continuous(range = c(3, 15)) +  
  scale_color_manual(values = c("blue", "pink")) +  
  labs(
    title = "High School Sports Participation by Sport and Gender",
    x = "Number of Schools Offering the Sport",
    y = "Number of Participants",
    size = "Participants",
    color = "Gender",
    caption = "Data: NFHS"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.key.size = unit(1.5, "lines"),  
    plot.margin = margin(t = 10, r = 30, b = 10, l = 10)  
  )



```

Notes:



--------
Little League Story:
- Participation data

Motion Graphic:
- 

Overview Story:
- Broad view of HS NIL policies across the country. 

Sponsorship Story:


Transfers:


Youngest NIL Athletes/TKD

NIL Contrarians:
- Viewpoints arguing against implementation of NIL policies for HS athletes. 

Collectives:
- YOKE: Software service that partners with Louisiana, a school can set up an NIL club where people on the team can sign up and produce content for money. 

Final Thoughts:
- Map of known HS NIL companies across the country (Cooper)




--------------

Meeting with Prof. Hyman on 11/21/24

Taking out states that don't have NIL policies (count up the total number of high school athletes)

- The way policies differ from one another (if md is the only state that prohibits video game NIL deals): Showing the collection of policies 
   - Decent amount of variation, how it's defined or the policy. Stack the policies as images, one on top of the other (four or five examples of here is what this state says)
   
   Penalty provisions, including the three-strike states, collectives
   
  Table: Here's a term, here's how this state defines it compared to another. How most states define it. How they are defining or choosing not to define something.
  
  
State association is to distance themselves from liability. Number 

All about differences, either explicit or gaps in ambiguity in the way policies are phrased.

Reader perspective: Some might not know NIL, some people don't know about sports. With regard to policies, what questions: Can a 15 year old kid do a commercial with a Smith and Wesson firearm? Address that curiosity.

NFHS data: 
- Participation data is relevant. 
