###Volleyball Z-Scores Analysis

_Step 1: Load in the Tidyverse and our 2024 match data._

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
matches24 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

_Step 2: Set up a code block to create team averages and z-scores (but change the metrics used in the tutorial)._

We'll first need to add up our totals for the stats we're using in this analysis.


```{r}
team_total_stats <- 
  matches24 |>
  group_by(team) |>
  summarise(
    sets = sum(s.x),
    tot_kills = sum(kills),
    tot_digs = sum(digs),
    tot_aces = sum(aces),
    avg_hit_pct = mean(hit_pct),
    tot_errors = sum(errors),
  ) |>
  select(team, sets, tot_kills, tot_digs, tot_aces, avg_hit_pct, tot_errors) 

```

Then, to find our averages, we need to divide our totals by the number of sets played throughout the season.

```{r}

vbstatspset <- 
  team_total_stats |>
 mutate(
    killsperset = tot_kills/sets,
    digsperset = tot_digs/sets,
    acesperset = tot_aces/sets,
    errorsperset = tot_errors/sets) |>
  select(team, killsperset, digsperset, acesperset, avg_hit_pct, errorsperset) 

```


Finally, we can calculate z-scores for our various stats.

```{r}

vbstatspset <- vbstatspset |>
   mutate(
   killsz = as.numeric(scale(killsperset, center = TRUE, scale = TRUE)),
   acesz = as.numeric(scale(acesperset, center = TRUE, scale = TRUE)),
   digsz = as.numeric(scale(digsperset, center = TRUE, scale = TRUE)),
   hitpctz = as.numeric(scale(avg_hit_pct, center = TRUE, scale = TRUE)),
   errorsz = as.numeric(scale(errorsperset, center = TRUE, scale = TRUE)),
  total_vbz = as.numeric(killsz + acesz + digsz + hitpctz + errorsz)
 )

head(vbstatspset)
```

Now we can take a look at our schools and see who our top performers are. 

```{r}

vbstatspset |>
  arrange(desc(total_vbz))

```

After creating a composite z-score, it appears that the school with the highest score is...Valparaiso with a z-score of 6.5518? Followed by Tennessee Tech, N.C. A&T, Georgia State, Wright State, Boise State, Saint Francis, Western Michigan, Bryant and Hofstra. 


So then where does Maryland fall on this list? Let's make a rank column to find out. 

```{r}

vbstatspset |>
arrange(desc(total_vbz)) |>  
  mutate(rank = row_number()) |>
  filter(team == "Maryland")
  
```

So we see that in our data set of 234 schools (available for analysis based on stats tracked), the Terps are ranked 196 in terms of composite z-score. However, a z-score of 0.6879 doesn't necessarily mean they're a "bad" team, but just that they're very average. Let's also take a look at the company the school keeps in the rankings.

```{r}

vbstatspset |>
arrange(desc(total_vbz)) |>  
  mutate(rank = row_number()) |>
  filter(rank >= 190 & rank <= 200)

```

Schools that are similarly ranked include Gonzaga, High Point, Tennessee State, Penn State, Houston, UC Riverside, Illinois State, SFA and Southern University.

Story ideas? 

My big question here is what is a 9-5 team from the Missouri Valley Conference (Valparaiso) doing so successfully to be this high on the list? Same with a lot of these mid-major schools in the top 10.

What are the keys to Georgia Southern, with a z-score of over 2.6 for digs, being elite at generating digs?

What can we find out about the Pittsburgh Panthers outperforming top schools in hitting percentage? And why is that team struggling in terms of digs? If they shored that issue up, what's their ceiling?

-----------------
Volleyball Z-Score Writeup

```{r}

vbstatspset |>
arrange(avg_hit_pct) |>
  mutate(rank = row_number()) |>
  filter(team == "Western Ill.")
 
```


Introducing the most gritty women's volleyball team in the country---The Western Illinois University Leathernecks. 

What does a 6-5 mid-major school from the Missouri Valley Conference bring to the table to be deserving of such a title? They find a way to generate digs at an elite level. So much  that they lead all Division I schools in digs per set. To put it in context, the Leathernecks nearly five more digs per set the Nebraska Cornhuskers and nearly six more than the Pittsburgh Panthers, who are currently the top two schools in the country.

The conditions a ripe for Western Illinois to take the crown for this stat. A high rate of digs across their games can obviously tell us they know what they're doing in that regard. However, it also tells us they might be on the backfoot during points and compensating through digs. The Leathernecks are poor at generating quick points through aces, holding the 11th worst ranked metrics for aces among 343 Division I schools, and are ranked No. 239 in average hitting percentage.

Their two tallest players are also six feet and two inches tall, with a majority of the roster falling between five feet and five inches to five feet and ten inches. This suggests that they might not be the most equipped roster to generate kills or blocks either.

However, all of those figures help clear up this picture. This is a team that struggles to close out points quickly, but is not afraid to do the dirty work in diving for balls during points. They refuse to count themselves down and out during these points, and that shows a lot of grit.



