
#Presentation 2: NBA Last Two-Minute Report

GitHub Link to Final Presentation: https://github.com/pabs-s/jour479x_fall_2024/blob/main/presentations/second_presentation.Rmd

This presentation focuses on the “NBA Officiating Last Two Minute Reports,” which are publicly accessible tables that share information related to the accuracy of calls made in the last two minutes of each game. Using this information, I'd like to identify the teams and players that have been either the beneficiaries or victims of incorrect officiating decisions. I'd also like to assess (to a degree) the accuracy of NBA referees in these late-game situations when calls have even more impact.

In my background research for this presentation, I found three examples of others who have attempted a similar analysis. The first source, which is from a GitHub user named atlhawksfanatic, offers a dataset formatted for R that is updated through Nov. 8, 2024. Here is the link to the GitHub page (https://github.com/atlhawksfanatic/L2M?tab=readme-ov-file).

As shown below, this data currently contains 83,987 rows and 64 columns. The time range begins on March 1, 2015 and ends on Nov. 8, 2024. It contains references to all 30 NBA teams. Among those 64 columns are data such as period (Q4 and Q5), time of the call, call type, players committing the foul/initiating the call and players disadvantaged in the call. There are also columns for game dates, official names, home and away teams, as well as ID signifiers for officials, games and teams, and scores.   

It's worth noting that there are additional time columns, one for minutes and another for seconds. I can also utilize columns for the network a game is shown on and whether it was shown on national TV. 


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

library(tidyverse)
library(ggplot2)
library(ggrepel)
library(janitor)
library(lubridate)
library(ggalt)
library(cowplot)
library(dplyr)
library(scales)
library(gt)
library(knitr)

```

```{r}

l2m <- read_csv("https://raw.githubusercontent.com/pabs-s/jour479x_fall_2024/refs/heads/main/presentations/l2m.csv")
colnames(l2m)

```

###Question 1: What are the most common types of incorrect calls? 
Subcategories for this question include the accuracy of shooting fouls and percentage of overall incorrect calls in the dataframe. 

```{r}

incorrect_calls <- l2m |>
  filter(decision %in% c("INC", "IC")) |>
  group_by(call_type) |>
  summarise(count = n()) |>
  arrange(desc(count))

top_ten_inc <- incorrect_calls |>
  slice_head(n = 10)  

ggplot(top_ten_inc, aes(x = reorder(call_type, count), y = count)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  scale_y_continuous(breaks = seq(0, max(top_ten_inc$count), by = 100)) +
  labs(
    title = "Shooting Fouls Spell End of Game Trouble for NBA Refs",
    x = "Call Type",
    y = "Count of Incorrect Calls",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez") +
  theme_minimal()

```

In looking at the results of this question, it's clear that NBA referees struggle the most with incorrectly calling shooting fouls in end-of-game scenarios. In fact, four of the top five most incorrectly called plays are fouls, with the only exception being traveling violations.

But I want to contextualize how often these shooting fouls are incorrectly called.

```{r}

shooting_fouls <- l2m |>
  filter(call_type == "Foul: Shooting")  

shootfoul_count <- shooting_fouls |>
  group_by(decision) |>        
  summarise(count = n())

ggplot(shootfoul_count, aes(x = decision, y = count, fill = decision)) +
  geom_bar(stat = "identity") +
  labs(
    title = "NBA Refs Are Overwhelmingly Accurate with Shooting Foul Calls",
    x = "Decision Type",
    y = "Count",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez"
  ) +
  theme_minimal()

```

When looking at all of the instances of fourth quarter and overtime shooting fouls, it's apparent that NBA referees overwhelmingly call (or don't call) shooting fouls correctly. In fact, of the 21,005 shooting fouls listed in the reports, only roughly 6.6% of them are considered incorrectly called or non-called.


```{r}

total_shooting_fouls <- nrow(shooting_fouls)

inc_sf <- shooting_fouls |>
  filter(decision %in% c("IC", "INC")) |>  
  nrow()

pct_inc_sf <- (inc_sf / total_shooting_fouls) * 100

cat("Percentage of IC + INC out of total shooting fouls:", pct_inc_sf, "%\n")

```

Now, let's see the overall percentage of incorrect calls throughout the entire L2M dataframe.

```{r}

sum(incorrect_calls$count, na.rm = TRUE) #5912
#83,987

(5912/83987) * 100 

```

The result is that in last 2-minute situations NBA referees get roughly 7% of calls incorrect. That is a good number to keep in mind moving through the rest of this analysis.

###Question 2: 
Who are the players that have been either the beneficiaries or victims of incorrect officiating decisions?

```{r}

inc_ic_calls <- l2m |>
  filter(decision %in% c("INC", "IC"))  

committing_counts <- inc_ic_calls |>
  group_by(committing) |>
  summarise(com_count = n(), .groups = "drop")  

disadvantaged_counts <- inc_ic_calls |>
  group_by(disadvantaged) |>
  summarise(dis_count = n(), .groups = "drop")

player_counts <- full_join(
  committing_counts, disadvantaged_counts,
  by = c("committing" = "disadvantaged")) |>
  rename(player = committing) 

player_counts <- player_counts |>
  mutate(
    dis_count = replace_na(dis_count, 0),
    com_count = replace_na(com_count, 0))

```

```{r}

player_counts <- player_counts |>
  mutate(total_count = dis_count + com_count) |>
  arrange(desc(total_count))

top_15_players <- player_counts |>
  filter(!is.na(player)) |> 
  arrange(desc(total_count)) |> 
  slice(1:15)

```

```{r}

ggplot(top_15_players, aes(x = dis_count, xend = com_count, y = reorder(player, total_count))) +
  geom_dumbbell(
    colour = "grey", 
    size = 2,         
    dot_guide = TRUE, 
    dot_guide_size = 0.5) +
  geom_point(aes(x = dis_count), color = "blue", size = 3) +  
  geom_point(aes(x = com_count), color = "red", size = 3) +   
  labs(
    title = "Top 15 Players: Disadvantaged vs. Committing Counts",
    subtitle = "Involvement in incorrect calls and no calls in NBA L2M report.",
    x = "Number of Plays",
    y = "Player",
    caption = "Blue = Disadvantaged, Red = Committing | Data: atlhawksfanatic (GitHub) | By Pablo Suarez") +
  theme_minimal()

```

Above is a list of the 15 players with the highest combined counts of disadvantaged and committed calls. A previous version of this chart showed that Al Horford, one of the longest standing NBA veterans, was very much an outlier regarding how involved he is in these calls. After further review, it turns out that an error in my code resulted in his figures being miscounted. I knew this also had to be a mistake because of the disparity with other long-tenured players in the graphic: LeBron James (22 years), Kyle Lowry (19 years), Paul George (14 years) and Russell Westbrook (17 years).

After fixing the graphic, Nikola Jokic leads all NBA players for references in plays with incorrect calls and no calls with a count of 117. He's also the only player in the top 8 that has been the disadvantaged player more than the committing player.

Conversely, players like Horford, Draymond Green and Karl-Anthony Towns are major offenders in terms of having high disparity between the instances where they are disadvantaged compared to committing fouls or turnovers.

Roughly half of these players are considered centers or power forwards (a.k.a., big men). This might be because their position on the court would make them susceptible to contact. The players who would be considered exceptions to this rule, like James Harden, Westbrook and Jimmy Butler, have also made a living by frequently getting into the paint for layups. 

Looking more into Nikola Jokic, I isolated the 2-minute report data to only show instances where he was referenced, which resulted in 1,562 plays. Filtering that data even further, I pulled all of the plays involving Jokic where the call was either incorrectly called or not called. That narrowed the dataframe down to 117 plays. That means that out of all of the last 2-minute report plays involving Jokic, roughly 7.5% of them ended with an incorrect call by the officials. Slightly higher than the league baseline that was previously established.

```{r}
# code for checking horford disparity 
# horford <- l2m |> filter(disadvantaged == "Al Horford" | committing == "Al Horford")
# horford |> filter(decision %in% c("INC", "IC")) |> group_by(committing) |> summarise(com_count = n(), .groups = "drop")
# horford |> filter(decision %in% c("INC", "IC")) |> group_by(disadvantaged) |> summarise(com_count = n(), .groups = "drop")

jokic <- l2m |> filter(disadvantaged == "Nikola Jokic" | committing == "Nikola Jokic")
jokic_inc_calls <- jokic |> filter(decision %in% c("INC", "IC")) 

```

The next codeblock helps create a distribution of 117 incorrect calls Jokic is involved in. 

```{r}

jokic_count <- jokic_inc_calls |>
  group_by(call_type, decision) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count))

jokic1 <- ggplot(jokic_count, aes(x = call_type, y = count, fill = decision)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("INC" = "blue", "IC" = "red")) +
  labs(
    title = "Jokic Gets a Good Whistle with Offensive Fouls",
    x = "Call Type",
    y = "Count",
    fill = "Decision",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(10, 10, 10, 30))


```

So what about a distribution of his committing versus disadvantaged fouls?

```{r}

jokic_call_summary <- jokic_inc_calls |>
  filter(committing == "Nikola Jokic" | disadvantaged == "Nikola Jokic") |> 
  mutate(role = case_when(
    committing == "Nikola Jokic" ~ "Committing",
    disadvantaged == "Nikola Jokic" ~ "Disadvantaged")) |> 
  group_by(call_type, role) |> 
  summarise(count = n()) |>
  arrange(desc(count))

jokic2 <- ggplot(jokic_call_summary, aes(x = call_type, y = count, fill = role)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("Disadvantaged" = "blue", "Committing" = "red")) +
  labs(
    x = "Call Type",
    fill = "Role") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(jokic1, jokic2)


```
It appears that the incorrect calls Jokic is predominantly involved in are mostly fouls, but specifically personal, shooting and loose ball fouls. It also seems like he loves to initiate contact, and he often gets away with more than he probably should when it comes to offensive fouls. 

However, that gets balanced out by him being on the unfortunate side of calls for personal and shooting fouls. Regardless, it just seems like contact always finds him.

### Question 3: What is the distribution of incorrect calls by year? 
Other findings in this section include the number of incorrect calls between the 4Q versus OT 

In this next section, I'm interested in finding out whether late game refereeing has changed significantly over the years since these reports were introduced. 

```{r}

l2m_summary <- l2m |>
  filter(decision %in% c("IC", "INC", "CC", "CNC")) |> 
  filter(season >= 2015 & season <= 2024) |> 
  group_by(season, decision) |> 
  summarise(count = n())

ggplot(l2m_summary, aes(x = season, y = count, color = decision, group = decision)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "NBA Late-Game Refereeing is Consistent Except for Correct No Calls",
    subtitle = "Correct and incorrect calls by year.",
    x = "Season",
    y = "Number of Calls",
    color = "Decision Type",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez"
  ) +
  scale_color_manual(
    values = c("CC" = "green", "CNC" = "blue", "IC" = "red", "INC" = "orange"),
    labels = c("CC" = "Correct Call", 
               "CNC" = "Correct No Call", 
               "IC" = "Incorrect Call", 
               "INC" = "Incorrect No Call")
  ) +
  scale_x_continuous(breaks = seq(2015, 2025, by = 1)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")


```
Overall, it appears that NBA referees seem to have a decent handle on accurately calling games. Only the plays classified as correct no calls seem to fluctuate from season to season. I would think that no call plays are just dependent on how much action there is at the end of games. Closer games would typically require more officiating compared to blowouts or decisive victories since the team that's ahead would be running out the clock. 

Next, let's see how calls between the last two minutes of the fourth quarter compare to overtime plays. 

```{r}

q4ot_counts <- l2m |>
  filter(decision %in% c("IC", "INC", "CC", "CNC")) |> 
  group_by(period, decision) |>
  summarise(count = n()) |> 
  arrange(period, desc(count))

ggplot(q4ot_counts, aes(x = decision, y = count, fill = period)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c("Q4" = "skyblue", "Q5" = "coral"),
    labels = c("Q4" = "4th Quarter", "Q5" = "Overtime")
  ) +
  labs(
    title = "Comparison of Call Decisions in Q4 and Q5",
    x = "Decision Type",
    y = "Count",
    fill = "Period"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    legend.position = "bottom")

```
I should've seen that one coming. Obviously there's a significant difference between Q4 and OT calls as not every games goes to OT. 

Instead, let's see what refs are calling in OT plays. 

```{r}

q5_calls <- l2m |>
  filter(period == "Q5", !is.na(call_type)) |> 
  group_by(call_type) |> 
  summarise(count = n()) |> 
  arrange(desc(count)) |> 
  slice_head(n = 10)

ggplot(q5_calls, aes(x = reorder(call_type, count), y = count)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() + 
  labs(
    title = "Common Fouls Are Common in Overtime Scenarios",
    x = "Call Type",
    y = "Count",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez"
  ) +
  theme_minimal()

```
Again, nothing too crazy compared to what we've already seen. Like other charts, it appears that fouls take the cake in overtime as well. 

```{r}

all_ot_calls <- l2m |> filter(period == "Q5", !is.na(call_type)) 

tot_ot_calls <- nrow(all_ot_calls) #9900
  
inc_ot_calls <- all_ot_calls |> filter(decision %in% c("IC", "INC")) |> nrow() #752

pct_inc_ot_calls <- (inc_ot_calls / tot_ot_calls) * 100

cat("Percentage of IC + INC OT calls out of total number of OT calls:", pct_inc_ot_calls, "%\n")

```

After filtering the two minute report dataframe for only overtime calls, I calculated the percentage of incorrectly called or uncalled fouls and turnovers. There were 9,900 total plays, with 752 of those listed as incorrect calls and no calls. With some simple math, I found that the percentage is roughly 7.6% which is a marginally higher difference than the 7% baseline. 

```{r}

filtered_ot_calls <- all_ot_calls |>
  filter(call_type %in% c(
    "Foul: Personal",
    "Foul: Shooting",
    "Foul: Offensive",
    "Foul: Loose Ball",
    "Turnover: Traveling",
    "Foul: Defensive 3 Second",
    "Instant Replay: Support Ruling",
    "Stoppage: Out-of-Bounds",
    "Foul: Personal Take",
    "Turnover: 24 Second Violation"
  )) |>
  mutate(call_type = case_when(
    call_type == "Turnover: 24 Second Violation" ~ "24 Sec. Violation",
    call_type == "Instant Replay: Support Ruling" ~ "Instant Replay",
    TRUE ~ call_type 
  ))

ggplot(filtered_ot_calls, aes(x = decision, fill = decision)) +
  geom_bar() +
  facet_wrap(~call_type, scales = "free_y") +  
  theme_minimal() +
  labs(
    title = "There's an OT Traveling Problem",
    subtitle = "Distribution of top ten calls in NBA overtimes since March 2015.",
    x = "Decision",
    y = "Count",
    fill = "Decision",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10)
  )

```
Although we've established that NBA refs are still pretty good compared to the 7% baseline. I was still interested in seeing the distribution of those top ten common calls. I can't say that the results are entirely surprising, but there are a few things worth noting.

For starters, instant replay calls were essentially entirely accurate. That's great news for NBA refs that when they leverage technology and take a close look, they are found to make the right calls all of the time. On the flip side, traveling violations seem to cause them a slew of problems. Granted travels account for a much smaller percentage of calls in the reports, they have by a significant margin the highest proportion of incorrect no calls compared to correct calls and no calls.

The "NA" values in the traveling section of the bar charts represent calls that were only "observable in enhanced video" by the referees who reviewed the play for the report. That signifies that between the 2015 season and now, there have been nearly 125 instances where referees failed to recognize a travel in overtime scenarios. 

Again, that's pretty impressive all considered, especially because actions occur so quickly on the court. It would be a fun exercise to look more in-depth at these calls to see how costly each one is and how it might've impacted games and teams. That's another challenge for another day.

###Question 4: Do games on national TV lead to more or less accurate calls at the end of games?

Considering that the NBA last 2-minute reports note whether a game is played on national TV and that it specifies the network if it is, I feel that it's worth assessing referee performance in games where "everybody" is watching. 

```{r}

nat_tv <- l2m |>
  filter(national_tv != "no")

tot_nat_tv <- nrow(nat_tv) 

inc_nattv_calls <- nat_tv |> 
  filter(decision %in% c("IC", "INC")) |> 
  nrow()

pct_inc_nattv_calls <- (inc_nattv_calls / tot_nat_tv) * 100

cat("Percentage of IC + INC calls out of total number of calls in national TV games:", pct_inc_nattv_calls, "%\n")

```

The short answer here is that NBA refs in nationally televised games are slightly better than the 7% baseline average. It's not a significant difference, but it's also not nothing. Intuitively, it makes sense. The NBA has a vested interest in ensuring that games shown nationally are adequately officiated. That's not to say other games aren't, but more viewers equals more scrutiny and the league is better off saving its slightly more accurate crews for these games.

Let's then take a look at the difference among networks. The reports specify three national networks---ESPN, TNT and NBA TV. 

```{r}

espn <- l2m |>
  filter(national_tv == "ESPN")

tot_espn <- nrow(espn)

inc_espn <- espn |> 
  filter(decision %in% c("IC", "INC")) |> 
  nrow()

pct_espn <- (inc_espn / tot_espn) * 100

cat("Percentage of IC + INC calls out of total number of calls in ESPN games:", pct_espn, "%\n")

```

```{r}

tnt <- l2m |>
  filter(national_tv == "TNT")

tot_tnt <- nrow(tnt)

inc_tnt <- tnt |> 
  filter(decision %in% c("IC", "INC")) |> 
  nrow()

pct_tnt <- (inc_tnt / tot_tnt) * 100

cat("Percentage of IC + INC calls out of total number of calls in TNT games:", pct_tnt, "%\n")

```

```{r}

nbatv <- l2m |>
  filter(national_tv == "NBATV")

tot_nbatv <- nrow(nbatv)

inc_nbatv <- nbatv |> 
  filter(decision %in% c("IC", "INC")) |> 
  nrow()

pct_nbatv <- (inc_nbatv / tot_nbatv) * 100

cat("Percentage of IC + INC calls out of total number of calls in NBA TV games:", pct_nbatv, "%\n")

```

Referee performance across ESPN and TNT games is relatively normal. It's the NBA TV games that see a significant increase in the percentage of incorrectly called and uncalled plays, a roughly 5% increase to be more precise. That seems unusual compared to what I've observed up to this point.  

Let's see what we can find out. First I'm taking a look at what's predominantly called in these NBA TV games.

```{r}

nbatv_count <- l2m |>
  filter(national_tv == "NBATV") |> 
  group_by(call_type) |> 
  summarise(count = n()) |> 
  arrange(desc(count)) |> 
  slice_head(n = 10)

ggplot(nbatv_count, aes(x = reorder(call_type, count), y = count)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() + 
  labs(
    title = "Fouls Are A Dime a Dozen in NBA TV Games",
    x = "Call Type",
    y = "Count",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez"
  ) +
  theme_minimal()


```
Again, nothing revolutionary here. Common fouls seem to be what draws whistles. 

```{r}

filtered_nbatv <- nbatv |>
  filter(call_type %in% c(
    "Foul: Shooting",
    "Foul: Personal",
    "Foul: Offensive",
    "Foul: Loose Ball",
    "Instant Replay: Support Ruling",
    "Turnover: Traveling",
    "Turnover: 24 Second Violation",
    "Turnover: 5 Second Inbound",
    "Foul: Personal Take",
    "Foul: Defensive 3 Second"
  )) |>
  mutate(call_type = case_when(
    call_type == "Turnover: 24 Second Violation" ~ "24 Sec. Violation",
    call_type == "Instant Replay: Support Ruling" ~ "Instant Replay",
    call_type == "Turnover: 5 Second Inbound" ~ "5 Sec. Inbound",
    TRUE ~ call_type 
  ))

ggplot(filtered_nbatv, aes(x = decision, fill = decision)) +
  geom_bar() +
  facet_wrap(~call_type, scales = "free_y") +  
  theme_minimal() +
  labs(
    title = "Traveling Violations are Also a Problem Here",
    subtitle = "Distribution of top ten calls in NBA TV games since March 2015.",
    x = "Decision",
    y = "Count",
    fill = "Decision",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10)
  )

```
Similar to what I saw with the overtime calls, there seems to be an issue with calling traveling violations in these NBA TV games. The 5-second inbound calls aren't entirely as worrisome as they appear. There are only five total instances across the NBA TV dataframe I made and four of them happened in 2015, while the remaining one occurred in 2016. 

I would reason that the cause behind the higher percentage of incorrect calls is due to having a lower number of calls to examine, which would naturally skew percentages. 

### Question 5: What does call accuracy look like when game scores are within two points?

```{r}

close_games <- l2m |>
  filter(abs(home_score - away_score) <= 2)

close_games

```

```{r}

tot_close <- nrow(close_games)

inc_close <- close_games |> 
  filter(decision %in% c("IC", "INC")) |> 
  nrow()

pct_close <- (inc_close / tot_close) * 100

cat("Percentage of IC + INC calls out of total number of calls in two-point games:", pct_close, "%\n")


```


Question 6: Who are the teams that have been either the beneficiaries or victims of incorrect officiating decisions?

```{r}

teams <- c("Celtics", "Nets", "Knicks", "76ers", "Raptors", 
           "Bulls", "Cavaliers", "Pistons", "Pacers", "Bucks", 
           "Hawks", "Hornets", "Heat", "Magic", "Wizards", 
           "Nuggets", "Timberwolves", "Thunder", "Trail Blazers", 
           "Jazz", "Warriors", "Clippers", "Lakers", "Suns", 
           "Kings", "Mavericks", "Rockets", "Grizzlies", 
           "Pelicans", "Spurs")

top_teams <- player_counts |>
  filter(!is.na(player), player %in% teams) |>  
  arrange(desc(total_count)) |>  
  slice(1:15)  

```

Let's now use the same analysis to identify the most disadvantaged teams when it comes to incorrect calls. 

```{r}

teams <- c("Celtics", "Nets", "Knicks", "76ers", "Raptors", 
           "Bulls", "Cavaliers", "Pistons", "Pacers", "Bucks", 
           "Hawks", "Hornets", "Heat", "Magic", "Wizards", 
           "Nuggets", "Timberwolves", "Thunder", "Trail Blazers", 
           "Jazz", "Warriors", "Clippers", "Lakers", "Suns", 
           "Kings", "Mavericks", "Rockets", "Grizzlies", 
           "Pelicans", "Spurs")

top_teams <- player_counts |>
  filter(!is.na(player), player %in% teams) |>  
  arrange(desc(total_count)) |>  
  slice(1:15)  

ggplot(top_teams, aes(x = dis_count, xend = com_count, y = reorder(player, total_count))) +
  geom_dumbbell(
    colour = "grey", 
    size = 2,         
    dot_guide = TRUE, 
    dot_guide_size = 0.5) +
  geom_point(aes(x = dis_count), color = "blue", size = 3) +  
  geom_point(aes(x = com_count), color = "red", size = 3) +   
  labs(
    title = "Top 15 Teams: Disadvantaged vs. Committing Counts",
    subtitle = "Involvement in incorrect calls and no calls in NBA L2M report.",
    x = "Number of Plays",
    y = "Player",
    caption = "Blue = Disadvantaged, Red = Committing | Data: atlhawksfanatic (GitHub) | By Pablo Suarez") +
  theme_minimal()

```
It's deja vu for Jokic and his Nuggets. After following the same process to identify the top 15 teams and their involvement in incorrect calls and no calls, the Denver Nuggets lead the league with 53 disadvantaged calls. 

It's also interesting that the Memphis Grizzlies lead all teams with five committing calls. That's possibly because there are many more possibilities for a call to be disadvantageous for a team rather than a team collectively commit some sort of offense.  

For context, the graphic above is strictly for calls that are attributed directly to the team in the reports. Getting an even clearer picture of the frequency for a team's advantageous or disadvantageous calls requires totaling all instances of referenced players. For what it's worth. I think it's possible to figure out moving forward, as the "comments" column in the dataframe contains abbreviations for referenced players. Here's an example involving Jokic himself, "Jokic (DEN) establishes himself in a screening position..." I would filter the column for those team abbreviations and then group those rows based on the committing vs disadvantaged result.

Okay, tangent over, now back to the Nuggets. Here is a distribution of those calls involving the team.

```{r}

nuggets <- l2m |> filter(disadvantaged == "Nuggets" | committing == "Nuggets")
nuggets_inc_calls <- nuggets |> filter(decision %in% c("INC", "IC")) 

nuggets_count <- nuggets_inc_calls |>
  group_by(call_type, decision) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count))

nuggs1 <- ggplot(nuggets_count, aes(x = call_type, y = count, fill = decision)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("INC" = "blue", "IC" = "red")) +
  labs(
    title = "Nuggets Suffer from Uncalled Opponent Travels",
    x = "Call Type",
    y = "Count",
    fill = "Decision",
    caption = "Data: atlhawksfanatic (GitHub) | By Pablo Suarez") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(10, 10, 10, 30))


```


```{r}

nuggs_call_summary <- nuggets_inc_calls |>
  filter(committing == "Nuggets" | disadvantaged == "Nuggets") |> 
  mutate(role = case_when(
    committing == "Nuggets" ~ "Committing",
    disadvantaged == "Nuggets" ~ "Disadvantaged")) |> 
  group_by(call_type, role) |> 
  summarise(count = n()) |>
  arrange(desc(count))

nuggs2 <- ggplot(nuggs_call_summary, aes(x = call_type, y = count, fill = role)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("Disadvantaged" = "blue", "Committing" = "red")) +
  labs(
    x = "Call Type",
    fill = "Role") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(nuggs1, nuggs2)


```




----------------



Further Analysis:
- Who are the teams that have been either the beneficiaries or victims of incorrect officiating decisions? Is there a difference in home versus away calls for some of them?
Given that I've already done this for players, I think I should be able to follow a similar process for the teams. I'll start by creating one big filter that includes each team's name, but it shouldn't give me too much trouble. 

- Which officials have the highest amount of incorrect calls? Additionally, what's the percentage of incorrect calls made by these officials?
This should also be possible as there are three separate columns to denote the officials for each game. This would be an approximate value as there is no data that suggests which official initiated a call. But I think this should be relatively straightforward to accomplish.


